# Linear (Gaussian) Models

{{< include shared-config.qmd >}}

:::{.callout-note}
This content is adapted from Dobson & Barnett, An Introduction to Generalized Linear Models, 4th edition, Chapters 2-6\]
:::

```{r setup, include = FALSE}
options('digits' = 4)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
pander::panderOptions("table.emphasize.rownames", FALSE)

```

Functions from these packages will be used throughout this document:

```{r}
library(pander) # format tables for markdown
library(ggplot2) # graphics
library(ggeasy) # help with graphics
library(dplyr) # manipulate data
library(haven) # import Stata files
library(knitr) # format R output for markdown
library(tidyr) # Tools to help to create tidy data
library(plotly) # interactive graphics
library(dobson) # datasets from Dobson and Barnett 2018
library(parameters) # format model output tables for markdown
library(conflicted) # check for conflicting function definitions
```

## Understanding Gaussian Linear Regression Models

### Motivating example: birthweights and gestational age

Suppose we want to learn about the distributions of birthweights for (human) babies born at different gestational ages and with different chromosomal sexes (Dobson and Barnett, Example 2.2.2):

```{r}

data("birthweight", package = "dobson")

bw = 
  birthweight |> 
  pivot_longer(
    cols = everything(),
    names_to = c("sex", ".value"),
    names_sep = "s "
  ) |> 
  mutate(
    sex = ifelse(sex == "boy", "male", "female"),
    male = (sex == "male") |> as.integer())  |> 
  rename(age = `gestational age`)
```

```{r}

plot1 = bw |> 
  ggplot(aes(
    x = age, 
    y = weight,
    linetype = sex,
    shape = sex,
    col = sex))  +
  theme_bw() +
  xlab("Gestational age (weeks)") +
  ylab("Birthweight (grams)") +
  # expand_limits(y = 0, x = 0) +
  geom_point(alpha = .7)

```

::: {.content-visible when-format="html"}
```{r}

ggplotly(plot1 + facet_wrap(~ sex))

```
:::

::: {.content-visible when-format="pdf"}
```{r}

print(plot1 + facet_wrap(~ sex))

```
:::

### Parallel lines regression

We don't have enough data to model the distribution of birth weight separately for each combination of gestational age and sex, so let's instead consider a (relatively) simple model for how that distribution varies with gestational age and sex.

#### Notation

Let:

-   $Y$ represent birthweight (measured in grams)
-   $X_1$ represent chromosomal sex:
    -   $X_1 = 0$ if female (XX)
    -   $X_1 = 1$ if male (XY)
-   $X_2$ represent gestational age at birth (measured in weeks).

::: callout-note
Female is the **reference level** for the categorical variable $X_1$ (chromosomal sex). The choice of a reference level is arbitrary and does not limit what we can do with the resulting model; it only makes it more computationally convenient to make inferences about comparisons involving that reference group.
:::

Now, consider the following model:

$$Y \sim N(\mu(X_1,X_2), \sigma^2)$$

$$\mu(X_1,X_2)\eqdef E[Y|X_1, X_2] = \beta_0 + \beta_1 X_1+ \beta_2 X_2$$

#### Implementing the Model in R

Here's how we can implement this model in R:

```{r}
bw_lm1 = lm(
  formula = weight ~ sex + age, 
  data = bw)

bw_lm1 |> 
  parameters(show_sigma = TRUE) |> 
  print_md()
```

Here's how this model looks, superimposed on the data:

```{r}
bw = 
  bw |> 
  mutate(`E[Y|X=x]` = fitted(bw_lm1)) |> 
  arrange(sex, age)

plot2 = 
  plot1 %+% bw +
  geom_line(aes(y = `E[Y|X=x]`))
```

::: {.content-visible when-format="pdf"}
```{r}
#| fig-cap: "Parallel-slopes model of birthweight"
print(plot2)

```
:::

::: {.content-visible when-format="html"}
```{r}
#| fig-cap: "Parallel-slopes model of birthweight"
ggplotly(plot2)

```
:::

#### Model assumptions and predictions

To learn what this model is assuming, let's plug in a few values.

According to this model, what's the mean birthweight for a female born at 36 weeks?

```{r}
pred_female = coef(bw_lm1)["(Intercept)"] + coef(bw_lm1)["age"]*36

# print(pred_female)
### built-in prediction: 
# predict(bw_lm1, newdata = tibble(sex = "female", age = 36))
```

$$E[Y|X_1 = 0, X_2 = 36] = \beta_0 + \beta_1 \cdot 0+ \beta_2 \cdot 36 =  `r pred_female`$$

What's the mean birthweight for a male born at 36 weeks?

```{r}
pred_male = 
  coef(bw_lm1)["(Intercept)"] + 
  coef(bw_lm1)["sexmale"] + 
  coef(bw_lm1)["age"]*36
```

$$
E[Y|X_1 = 1, X_2 = 36] = \beta_0 + \beta_1 \cdot 1+ \beta_2 \cdot 36 =  `r pred_male`
$$

What's the difference in mean birthweights between males born at 36 weeks and females born at 36 weeks?

$$
\begin{aligned}
& E[Y|X_1 = 1, X_2 = 36] - E[Y|X_1 = 0, X_2 = 36]\\
&= 
`r pred_male` - `r pred_female`\\
&=
`r pred_male - pred_female`
\end{aligned}
$$

Shortcut:

$$
\begin{aligned}
& E[Y|X_1 = 1, X_2 = 36] - 
E[Y|X_1 = 0, X_2 = 36]\\
&= (\beta_0 + \beta_1 \cdot 1+ \beta_2 \cdot 36) - (\beta_0 + \beta_1 \cdot 0+ \beta_2 \cdot 36) \\
&= \beta_1 \\ 
&=  `r coef(bw_lm1)["sexmale"]`
\end{aligned}
$$

Note that age doesn't show up in this difference: in other words, according to this model, the difference between females and males with the same gestational age is the same for every age.

That's an assumption of the model; it's built-in to the parametric structure, even before we plug in the estimated values of those parameters.

That's why the lines are parallel.

### Interactions

What if we don't like that parallel lines assumption?

Then we need to allow an "interaction" between age and sex:

$$
E[Y|X_1, X_2] = \beta_0 + \beta_1 X_1+ \beta_2 X_2 + \beta_3 (X_1 \cdot X_2)
$$

```{r}

bw_lm2 = lm(weight ~ sex + age + sex:age, data = bw)

```

Here are the estimated parameters ($\beta$s):

```{r}

bw_lm2 |> 
  parameters() |> 
  print_md()

```

Here's another way we could rewrite this model (by collecting terms involving $X_2$):

$$
E[Y|X_1, X_2] = \beta_0 + \beta_1 X_1+ (\beta_2 + \beta_3 X_1) X_2
$$

::: callout-note
If you want to understand a coefficient in a model with interactions, collect terms for the corresponding variable, and you will see what other variables are interacting with the variable you are interested in.
:::

In this case, the coefficient $X_2$ is interacting with $X_1$. So the slope of $Y$ with respect to $X_2$ depends on the value of $X_2$.

According to this model, there is no such thing as "*the* slope of birthweight with respect to age". There are two slopes, one for each sex.^[using the definite article "the" would mean there is only one slope.] We can only talk about "the slope of birthweight with respect to age among males" and "the slope of birthweight with respect to age among females".

Then: that coefficient is the difference in means per unit change in its corresponding coefficient, when the other collected variables are set to 0.

Here's how this model looks, superimposed on the data:

```{r}

bw = 
  bw |> 
  mutate(
    predlm2 = predict(bw_lm2)
  ) |> 
  arrange(sex, age)

plot1_interact = 
  plot1 %+% bw +
  geom_line(aes(y = predlm2))

```

::: {.content-visible when-format="pdf"}
```{r}
print(plot1_interact)
```
:::

::: {.content-visible when-format="html"}
```{r}
ggplotly(plot1_interact)

```
:::

Now we can see that the lines aren't parallel.

To learn what this model is assuming, let's plug in a few values.

According to this model, what's the mean birthweight for a female born at 36 weeks?

```{r}
pred_female = coef(bw_lm2)["(Intercept)"] + coef(bw_lm2)["age"]*36
```

$$E[Y|X_1 = 0, X_2 = 36] = \beta_0 + \beta_1 \cdot 0+ \beta_2 \cdot 36 + \beta_3 \cdot (0 * 36) =  `r pred_female`$$ What's the mean birthweight for a male born at 36 weeks?

```{r}
pred_male = 
  coef(bw_lm2)["(Intercept)"] + 
  coef(bw_lm2)["sexmale"] + 
  coef(bw_lm2)["age"]*36 + 
  coef(bw_lm2)["sexmale:age"] * 36
```

$$E[Y|X_1 = 0, X_2 = 36] = \beta_0 + \beta_1 \cdot 1+ \beta_2 \cdot 36 + \beta_3 \cdot 1 \cdot 36 =  `r pred_male`$$

What's the difference in mean birthweights between males born at 36 weeks and females born at 36 weeks?

$$
\begin{aligned}
& E[Y|X_1 = 1, X_2 = 36] - E[Y|X_1 = 0, X_2 = 36]\\ 
&= (\beta_0 + \beta_1 \cdot 1+ \beta_2 \cdot 36 + \beta_3 \cdot 1 \cdot 36)\\ 
&\ \ \ \ \  -(\beta_0 + \beta_1 \cdot 0+ \beta_2 \cdot 36 + \beta_3 \cdot 0 \cdot 36) \\
&= \beta_2 + \beta_3\cdot 36\\
&=  `r pred_male - pred_female`
\end{aligned}
$$

Note that age now does show up in the difference: in other words, according to this model, the difference in mean birthweights between females and males with the same gestational age can vary by gestational age.

That's how the lines in the graph ended up non-parallel.

### Stratified regression

We could re-write the interaction model as a stratified model, with a slope and intercept for each sex:

```{r}
bw_lm_strat = 
  bw |> 
  lm(
    formula = weight ~ sex + sex:age - 1, 
    data = _)

bw_lm_strat |> 
  parameters() |> 
  print_md()
```

### Curved-line regression

If we transform some of our covariates ($X$s) and plot the resulting model on the original covariate scale, we end up with curved regression lines:

```{r}

bw_lm3 = lm(weight ~ sex:log(age) - 1, data = bw)
library(palmerpenguins)

ggpenguins <- 
  palmerpenguins::penguins |> 
  dplyr::filter(species == "Adelie") |> 
  ggplot(
    aes(x = bill_length_mm , y = body_mass_g)) +
  geom_point() + 
  xlab("Bill length (mm)") + 
  ylab("Body mass (g)")

ggpenguins2 = ggpenguins +
  stat_smooth(
    method = "lm",
              formula = y ~ log(x),
              geom = "smooth") +
  xlab("Bill length (mm)") + 
  ylab("Body mass (g)")


```

::: {.content-visible when-format="pdf"}
```{r}
ggpenguins2 |> print()
```
:::

::: {.content-visible when-format="html"}
```{r}
ggpenguins2 |>  ggplotly()

```
:::

## Estimating Linear Models via Maximum Likelihood

### Review of one-sample inference

Previously, we learned how to fit outcome-only models of the form $p(X=x|\theta)$ to iid data $\mathbf x = (x_1,…,x_n)$ using maximum likelihood estimation:

$$\mathcal L(\mathbf x|\theta) = p(X_1 = x_1, …,X_n =x_n|\theta) = \prod_{i=1}^n p(X=x_i|\theta)$$

$$\ell(x|\theta) = \log{\mathcal L(x|\theta)}$$

$$\hat \theta_{ML} = \arg \max_\theta \ell(x|\theta)$$

We learned how to quantify our uncertainty about these maximum likelihood estimates; with sufficient sample size, $\hat \theta_{ML}$ has the approximate distribution:

$$
\hat\theta_{ML} \dot \sim N(\theta,\mathcal I(\theta)^{-1})
$$

For models in the "exponential family" of distributions, which includes the Gaussian, Poisson, Bernoulli, Binomial, Exponential, and Gamma distributions, $\mathcal I(\theta) = \text -E[\mathcal{l}''(X|\theta)]$, so we estimated $\mathcal I(\theta)$ using either $\mathcal I(\theta)|_{\theta = \hat \theta_{ML}}$ or $\mathcal{l}''(\mathbf x |\theta)|_{\theta = \hat \theta_{ML}}$.

Then an asymptotic approximation of a 95% confidence interval for $\theta_k$ is

$$\hat \theta_{ML} \pm z_{0.975} \times \left[\left(\hat{\mathcal I}(\hat \theta_{ML})\right)^{-1}\right]_{kk}$$

where $z_\beta$ the $\beta$ quantile of the standard Gaussian distribution.

### MLEs for Linear Regression

Let's use maximum likelihood again:

$$
\mathcal L(\mathbf y|\mathbf x,\beta, \sigma^2) = 
\prod_{i=1}^n (2\pi\sigma^2)^{-1/2} 
\exp{-\frac{1}{2\sigma^2}(y_i - x_i'\beta)^2}
$$

$$
\ell(\mathbf y|\mathbf x,\beta, \sigma^2) \propto -\frac{n}{2}\log{\sigma^2} - \frac{1}{2\sigma^2}\sum_{i=1}^n (y_i - x_i' \beta)^2
$$

$$
\ell'(\mathbf y|\mathbf x,\beta, \sigma^2) \propto -\frac{n}{2}\log{\sigma^2} - \frac{1}{2\sigma^2}\frac{d}{d\beta}\left(\sum_{i=1}^n (y_i - x_i' \beta)^2\right)
$$

A few tools from linear algebra will make this analysis go easier (see @fieller2018basics, Section 7.2 for details).

$$
f_{\beta}(\mathbf x) = (f_{\beta}(x_1), f_{\beta}(x_2), ..., f_{\beta}(x_n))^\top
$$

Let $\mathbf x$ and $\beta$ be vectors of length $p$, or in other words, matrices of length $p\times 1$:

$$
x = \begin{bmatrix}
x_{1} \\
x_{2} \\
\vdots \\
x_{p}
\end{bmatrix} \\
\beta = \begin{bmatrix}
\beta_{1} \\
\beta_{2} \\
\vdots \\
\beta_{p}
\end{bmatrix}
$$

Then

$$
x' \equiv x^\top \equiv [x_1, x_2, ..., x_p]
$$

and

$$
x'\beta = [x_1, x_2, ..., x_p] 
\begin{bmatrix}
\beta_{1} \\
\beta_{2} \\
\vdots \\
\beta_{p}
\end{bmatrix} = 
x_1\beta_1+x_2\beta_2 +...+x_p \beta_p 
$$

If $f(\beta)$ is a function that takes $\beta$ as input and outputs a scalar, such as $f(\beta) = x'\beta$, then:

$$
\frac{d}{d \beta} f(\beta)= 
\begin{bmatrix}
\frac{d}{d\beta_1}f(\beta) \\
\frac{d}{d\beta_2}f(\beta) \\
\vdots \\
\frac{d}{d\beta_p}f(\beta)
\end{bmatrix}
$$

In particular, if $f(\beta) = x'\beta$, then:

$$ 
\frac{d}{d \beta} x'\beta= 
\begin{bmatrix}
\frac{d}{d\beta_1}(x_1\beta_1+x_2\beta_2 +...+x_p \beta_p ) \\
\frac{d}{d\beta_2}(x_1\beta_1+x_2\beta_2 +...+x_p \beta_p ) \\
\vdots \\
\frac{d}{d\beta_p}(x_1\beta_1+x_2\beta_2 +...+x_p \beta_p )
\end{bmatrix} 
=
\begin{bmatrix}
x_{1} \\
x_{2} \\
\vdots \\
x_{p}
\end{bmatrix}
= \mathbf x
$$

In general:

$$
\frac{d}{d\beta} x'\beta = x
$$

This looks a lot like non-vector calculus, except that you have to transpose the coefficient.

Similarly,

$$
\frac{d}{d\beta} \beta'\beta = 2\beta
$$

This is like taking the derivative of $x^2$.

And finally, if $S$ is a $p\times p$ matrix, then:

$$
\frac{d}{d\beta} \beta'S\beta = 2S\beta
$$

Again, this is like taking the derivative of $cx^2$ with respect to $c$ in non-vector calculus.

Thus:

$$
\sum_{i=1}^n (y_i - f_\beta(x_i))^2 = (\mathbf y - X\beta)'(\mathbf y - X\beta)
$$

$$
(\mathbf y - X\beta)' = (\mathbf y' - (X\beta)') = (\mathbf y' - \beta'X')
$$

So

$$
\begin{aligned}
(\mathbf y - X\beta)'(\mathbf y - X\beta) &= (\mathbf y' - \beta'X')(\mathbf y - X\beta)\\
&= y'y - \beta'X'y - y'X\beta +\beta'X'X\beta\\
&= y'y - 2y'X\beta +\beta'X'X\beta
\end{aligned}
$$

So

$$
\begin{aligned}
\frac{d}{d\beta}\left(\sum_{i=1}^n (y_i - x_i' \beta)^2\right) &= 
\frac{d}{d\beta}(\mathbf y - X\beta)'(\mathbf y - X\beta)\\
&= \frac{d}{d\beta} (y'y - 2y'X\beta +\beta'X'X\beta)\\
&= (- 2X'y +2X'X\beta)
\end{aligned}
$$

So if $\ell(\beta,\sigma^2) =0$, then

$$
\begin{aligned}
0 &= (- 2X'y +2X'X\beta)\\
2X'y &= 2X'X\beta\\
X'y &= X'X\beta\\
(X'X)^{-1}X'y &= \beta
\end{aligned}
$$

The second derivative matrix $\ell_{\beta, \beta'} ''(\beta, \sigma^2;\mathbf X,\mathbf y)$ is negative definite at $\beta = (X'X)^{-1}X'y$, so $\hat \beta_{ML} = (X'X)^{-1}X'y$ is the MLE for $\beta$.

Similarly (not shown):

$$
\hat\sigma^2_{ML} = \frac{1}{n} (Y-X\hat\beta)'(Y-X\hat\beta)
$$

And

$$
\begin{aligned}
\mathcal I_{\beta} &= E[-\ell_{\beta, \beta'} ''(Y|X,\beta, \sigma^2)]\\
&= \frac{1}{\sigma^2}X'X
\end{aligned}
$$

So:

$$
Var(\hat \beta) \approx (\mathcal I_{\beta})^{-1} = \sigma^2 (X'X)^{-1}
$$

and

$$
\hat\beta \dot \sim N(\beta, \mathcal I_{\beta}^{-1})
$$ These are all results you have hopefully seen before, and in the Gaussian linear regression case they are exact, not just approximate.

In our model 2 above, this matrix is:

```{r}
bw_lm2 |> vcov()
```

Note that if we take the square roots of the diagonals, we get the standard errors listed in the model output:

```{r}

bw_lm2 |> vcov() |> diag() |> sqrt()

```

```{r}
bw_lm2 |> summary() |> coef() |> pander()
```

So we can do confidence intervals, hypothesis tests, and p-values exactly as in the one-variable case we looked at previously.

## Inference about Gaussian Linear Regression Models

Research question: is there really an interaction between sex and age?

$H_0: \beta_3 = 0$

$H_A: \beta_3 \neq 0$

$P(|\hat\beta_3| > |`r coef(bw_lm2)["sexmale:age"]`| \mid H_0)$ = ?

### Wald tests and CIs

R can give you Wald tests for single coefficients and corresponding CIs:

```{r "wald tests bw_lm2"}

bw_lm2 |> 
  parameters(ci_method = "wald") |> 
  print_md()
```

### One-parameter inference by hand

To understand what's happening, let's replicate these results by hand for the interaction term.

#### P-value

```{r}
beta_hat = coef(summary(bw_lm2))["sexmale:age", "Estimate"]
se_hat = coef(summary(bw_lm2))["sexmale:age", "Std. Error"]
dfresid = bw_lm2$df.residual
t_stat = abs(beta_hat)/se_hat
pval_t = pt(abs(t_stat), df = dfresid, lower = FALSE) * 2

```

$$
\begin{aligned}
&P\left(
| \hat \beta_3  | > 
| `r coef(bw_lm2)["sexmale:age"]`| \middle| H_0
\right) 
&= P\left(
\left| \frac{\hat\beta_3}{\hat{SE}(\hat\beta_3)} \right| > 
\left| \frac{`r beta_hat`}{`r se_hat`} \right| \middle| H_0
\right)\\ 
&= P\left(
| T_{`r dfresid`} | >  `r t_stat` \middle| H_0
\right)\\
&= `r pval_t`
\end{aligned}
$$ This matches the result in the table above.

#### Confidence interval

```{r}
confint_radius_t = se_hat * qt(p = 0.975, df = dfresid, lower = TRUE)
confint_t = beta_hat + c(-1,1) * confint_radius_t
print(confint_t)
```

This also matches.

### Gaussian approximations

Here are the asymptotic (Gaussian approximation) equivalents:

#### P-value

```{r}
pval_z = pnorm(abs(t_stat), lower = FALSE) * 2

print(pval_z)
```

#### Confidence interval

```{r}
confint_radius_z = se_hat * qnorm(0.975, lower = TRUE)
confint_z = 
  beta_hat + c(-1,1) * confint_radius_z
print(confint_z)
```

### Likelihood ratio statistics

```{r}

logLik(bw_lm2)
logLik(bw_lm1)

lLR = (logLik(bw_lm2) - logLik(bw_lm1)) |> as.numeric()
delta_df = (bw_lm1$df.residual - df.residual(bw_lm2))

d_lLR = function(x, df = delta_df) dchisq(x, df = df)

x_max = 1

chisq_plot = 
  ggplot() + 
  geom_function(fun = d_lLR) +
  stat_function( fun = d_lLR, xlim = c(lLR, x_max), geom = "area", fill = "gray") +
  geom_segment(aes(x = lLR, xend = lLR, y = 0, yend = d_lLR(lLR)), col = "red") + 
  xlim(0.0001,x_max) + 
  ylim(0,4) + 
  ylab("p(X=x)") + 
  xlab("log(likelihood ratio) statistic [x]") +
  theme_classic()
```

```{r}
pchisq(
  q = 2*lLR, 
  df = delta_df, 
  lower = FALSE)
```

::: {.content-visible when-format="pdf"}
```{r}
chisq_plot |> print()
```
:::

::: {.content-visible when-format="html"}
```{r}
chisq_plot |>  ggplotly()

```
:::

Now we can get the p-value:

```{r}
pchisq(2*lLR, df = delta_df, lower = FALSE)

```

### 

In practice you don't have to do this by hand; there are functions to do it for you:

```{r}

# built in
library(lmtest)
lrtest(bw_lm2, bw_lm1)

```

## Goodness of fit

### AIC and BIC

When we use likelihood ratio tests, we are comparing how well different models fit the data.

Likelihood ratio tests require "nested" models: one must be a special case of the other.

If we have non-nested models, we can instead use the Akaike Information Criterion (AIC) or Bayesian Information Criterion (BIC):

-   AIC = $-2 * \ell(\hat\theta) + 2 * p$

-   BIC = $-2 * \ell(\hat\theta) + p * \text{log}(n)$

where $\ell$ is the log-likelihood of the data evaluated using the parameter estimates $\hat\theta$, $p$ is the number of estimated parameters in the model (including $\hat\sigma^2$), and $n$ is the number of observations.

You can calculate these criteria using the `logLik()` function, or use the built-in R functions:

#### AIC in R

```{r}

-2 * logLik(bw_lm2) |> as.numeric() + 
  2*(length(coef(bw_lm2))+1) # sigma counts as a parameter here

AIC(bw_lm2)

```

#### BIC in R

```{r}

-2 * logLik(bw_lm2) |> as.numeric() + 
  (length(coef(bw_lm2))+1) * log(nobs(bw_lm2))

BIC(bw_lm2)

```

Large values of AIC and BIC are worse than small values. There are no hypothesis tests or p-values associated with these criteria.

### (Residual) Deviance

Let $q$ be the number of distinct covariate combinations in a data set.

```{r}
bw.X.unique = 
  bw |> 
  count(sex, age)

n_unique.bw  = nrow(bw.X.unique)
```

For example, in the `birthweight` data, there are $q = `r n_unique.bw`$ unique patterns:

```{r}

bw.X.unique |> 
  pander(
    row.names = rownames(bw.X.unique))
```

::: callout-note
If a given covariate pattern has more than one observation in a dataset, those observations are called **replicates**.
:::

Then the most complicated model we could fit would have one parameter (a mean) for each of these combinations, plus a variance parameter:

```{r}

lm_max = 
  bw |> 
  mutate(age = factor(age)) |> 
  lm(
    formula = weight ~ sex:age - 1, 
    data = _)

lm_max |> 
  parameters() |> 
  print_md()

```

We call this model the **full**, **maximal**, or **saturated** model for this dataset.

We can calculate the log-likelihood of this model as usual:

```{r}
logLik(lm_max)
```

We can compare this model to our other models using chi-square tests, as usual:

```{r}
lrtest(lm_max, bw_lm2)
```

The likelihood ratio statistic for this test is $$\lambda = 2 * (\ell_{\text{full}} - \ell) = `r 2 * (logLik(lm_max) - logLik(bw_lm2))`$$ where:

-   $\ell_{\text{max}}$ is the log-likelihood of the full model: `r logLik(lm_max)`
-   $\ell$ is the log-likelihood of our comparison model (two slopes, two intercepts): `r logLik(bw_lm2)`

This statistic is called the **deviance** or **residual deviance** for our two-slopes and two-intercepts model; it tells us how much the likelihood of that model deviates from the likelihood of the maximal model.

The corresponding p-value tells us whether there we have enough evidence to detect that our two-slopes, two-intercepts model is a worse fit for the data than the maximal model; in other words, it tells us if there's evidence that we missed any important patterns. (Remember, a nonsignificant p-value could mean that we didn't miss anything and a more complicated model is unnecessary, or it could mean we just don't have enough data to tell the difference between these models.)

### Null Deviance

Similarly, the *least* complicated model we could fit would have only one mean parameter, an intercept:

$$\text E[Y|X=x] = \beta_0$$ We can fit this model in R like so:

```{r}
lm0 = lm(weight ~ 1, data = bw)

lm0 |> parameters() |> print_md()

```

This model also has a likelihood:

```{r}
logLik(lm0)
```

And we can compare it to more complicated models using a likelihood ratio test:

```{r}

lrtest(bw_lm2, lm0)

```

The likelihood ratio statistic for the test comparing the null model to the maximal model is $$\lambda = 2 * (\ell_{\text{full}} - \ell_{0}) = `r 2 * (logLik(lm_max) - logLik(lm0))`$$ where:

-   $\ell_{\text{0}}$ is the log-likelihood of the null model: `r logLik(lm0)`
-   $\ell_{\text{full}}$ is the log-likelihood of the maximal model: `r logLik(lm_max)`

In R, this test is:

```{r}
lrtest(lm_max, lm0)
```

This log-likelihood ratio statistic is called the **null deviance**. It tells us whether we have enough data to detect a difference between the null and full models.

## Rescaling

### Rescale age {.smaller}

```{r}

bw = 
  bw |>
  mutate(
    `age - mean` = age - mean(age),
    `age - 36wks` = age - 36
  )

lm1c = lm(weight ~ sex + `age - 36wks`, data = bw)

lm2c = lm(weight ~ sex + `age - 36wks` + sex:`age - 36wks`, data = bw)

parameters(lm2c, ci_method = "wald") |> print_md()

```

Compare with what we got without rescaling:

```{r}
parameters(bw_lm2, ci_method = "wald") |> print_md()
```

## Prediction

### Prediction for linear models

$$
\begin{aligned}
\hat Y &= \hat E[Y|X=x] \\
&= x'\hat\beta \\
&= \hat\beta_0\cdot 1 + \hat\beta_1 x_1 + ... + \hat\beta_p x_p
\end{aligned}
$$

### prediction in R

```{r}

X = model.matrix(bw_lm1)
X |> as_tibble() |> print()

print(X[1,])
print(coef(bw_lm1))
print(X[1,] * coef(bw_lm1))
{X[1,] * coef(bw_lm1)} |> sum() |> print()

X %*% coef(bw_lm1) |> as.vector()

predict(bw_lm1)

predict(bw_lm1, se.fit = TRUE)

predict(bw_lm1, se.fit = TRUE)$fit
predict(bw_lm1, se.fit = TRUE)$se.fit

predict(bw_lm1, se.fit = TRUE, interval = "confidence")$fit |> as_tibble()

predict(bw_lm1, se.fit = TRUE, interval = "predict")$fit |> as_tibble()

```

The warning from the last command is: "predictions on current data refer to *future* responses" (since you already know what happened to the current data, and thus don't need to predict it). You could also supply `newdata` to get predictions for new combinations of predictors that you didn't see in your original dataset. See `?predict.lm` for more.

## Diagnostics

### Residuals

#### Definitions of residuals

-   Residuals: $$e_i = y_i - \hat E[Y|X=x]$$

-   For Gaussian models, $e_i$ can be seen as an estimate of $$\epsilon_i = Y_i - \text{E}[Y|X=x_i]$$

-   Standardized residuals: $$r_i = \frac{e_i}{\hat{SD}(e_i)}$$

-   For Gaussian models: $$\hat{SD}(e_i) \approx \frac{e_i}{\hat{\sigma}}$$

#### Characteristics of residuals

With enough data and a correct model, the residuals will be approximately Guassian distributed, with variance $\sigma^2$, which we can estimate using $\hat\sigma^2$: that is:

$$
e_i\ \dot \sim_{iid}\ N(0, \hat\sigma^2)
$$

Hence, with enough data and a correct model, the standardized residuals will be approximately standard Gaussian; that is,

$$
r_i\ \dot \sim_{iid}\ N(0,1)
$$

### Marginal distributions of residuals

To look for problems with our model, we can check whether the residuals $e_i$ and standardized residuals $r_i$ look like they have the distributions that they are supposed to have, according to the model.

First, we need to compute the residuals. R makes this easy:

#### (non-standardized) residuals in R

```{r}
resid(bw_lm2)
# check by hand
bw$weight - fitted(bw_lm2)
```

Success!

#### Standardized residuals in R

```{r}

rstandard(bw_lm2)
resid(bw_lm2)/sigma(bw_lm2)

```

These are not quite the same, because R is doing something more complicated and precise to get the standard errors. Let's not worry about those details for now; the difference is pretty small in this case:

```{r}

rstandard_compare_plot = 
  tibble(
    x = resid(bw_lm2)/sigma(bw_lm2), 
    y = rstandard(bw_lm2)) |> 
  ggplot(aes(x = x, y = y)) +
  geom_point() + 
  theme_bw() +
  coord_equal() + 
  xlab("resid(bw_lm2)/sigma(bw_lm2)") +
  ylab("rstandard(bw_lm2)") +
  geom_abline(
    aes(
    intercept = 0,
    slope = 1, 
    col = "x=y")) +
  labs(colour="") +
  scale_colour_manual(values="red")

```

::: {.content-visible when-format="html"}
```{r}
ggplotly(rstandard_compare_plot)
```
:::

::: {.content-visible when-format="pdf"}
```{r}
print(rstandard_compare_plot)
```
:::

Let's add these residuals to the `tibble` of our dataset:

```{r}

bw = 
  bw |> 
  mutate(
    fitted_lm2 = fitted(bw_lm2),
    
    resid_lm2 = resid(bw_lm2),
    # resid_lm2 = weight - fitted_lm2,
    
    std_resid_lm2 = rstandard(bw_lm2),
    # std_resid_lm2 = resid_lm2 / sigma(bw_lm2)
  )

bw |> 
  select(
    sex,
    age,
    weight,
    fitted_lm2,
    resid_lm2,
    std_resid_lm2
  )
```

Now let's build histograms:

#### Marginal distribution of (nonstandardized) residuals

```{r}

resid_marginal_hist = 
  bw |> 
  ggplot(aes(x = resid_lm2)) +
  geom_histogram()

```

::: {.content-visible when-format="html"}
```{r}
ggplotly(resid_marginal_hist)
```
:::

::: {.content-visible when-format="pdf"}
```{r}
print(resid_marginal_hist)
```
:::

Hard to tell with this small amount of data, but I'm a bit concerned that the histogram doesn't show a bell-curve shape.

##### Marginal distribution of standardized residuals

```{r}

std_resid_marginal_hist = 
  bw |> 
  ggplot(aes(x = std_resid_lm2)) +
  geom_histogram()

```

::: {.content-visible when-format="html"}
```{r}
ggplotly(std_resid_marginal_hist)
```
:::

::: {.content-visible when-format="pdf"}
```{r}
print(std_resid_marginal_hist)
```
:::

This looks similar, although the scale of the x-axis got narrower, because we divided by $\hat\sigma$ (roughly speaking).

Still hard to tell if the distribution is Gaussian.

### QQ plot of standardized residuals

Another way to assess normality is the QQ plot of the standardized residuals versus normal quantiles:

```{r}

library(ggfortify) 
# needed to make ggplot2::autoplot() work for `lm` objects

qqplot_lm2_auto = 
  bw_lm2 |> 
  autoplot(
    which = 2, # options are 1:6; can do multiple at once
    ncol = 1) +
  theme_classic()

```

::: {.content-visible when-format="html"}
```{r}
print(qqplot_lm2_auto)
```
:::

::: {.content-visible when-format="pdf"}
```{r}
print(qqplot_lm2_auto)
```
:::

If the Gaussian model were correct, these points should follow the dotted line.

::: callout-note
Fig 2.4 panel (c) in Dobson is a little different; they didn't specify how they produced it, but other statistical analysis systems do things differently from R.
:::

#### QQ plot - how it's built

Let's construct it by hand:

```{r}

bw = bw |> 
  mutate(
    p = (rank(std_resid_lm2) - 1/2)/n(), # "Blom's method"
    expected_quantiles_lm2 = qnorm(p)
  )

qqplot_lm2 = 
  bw |> 
  ggplot(
    aes(
      x = expected_quantiles_lm2, 
      y = std_resid_lm2, 
      col = sex, 
      shape = sex)
  ) + 
  geom_point() +
  theme_classic() +
  theme(legend.position='none') + # removing the plot legend
  ggtitle("Normal Q-Q") +
  xlab("Theoretical Quantiles") + 
  ylab("Standardized residuals")
```

We find the expected line like so:

```{r}
ps <- c(.25, .75)                  # reference probabilities
a <- quantile(rstandard(bw_lm2), ps)  # empirical quantiles
b <- qnorm(ps)                     # theoretical quantiles

qq_slope = diff(a)/diff(b)
qq_intcpt = a[1] - b[1] * qq_slope

qqplot_lm2 = 
  qqplot_lm2 +
  geom_abline(slope = qq_slope, intercept = qq_intcpt)

```

::: {.content-visible when-format="html"}
```{r}
ggplotly(qqplot_lm2)
```
:::

::: {.content-visible when-format="pdf"}
```{r}
print(qqplot_lm2)
```
:::

### Conditional distributions of residuals

If our Gaussian linear regression model is correct, the residuals $e_i$ and standardized residuals $r_i$ should have:

-   an approximately Gaussian distribution, with:
-   a mean of 0
-   a constant variance

This should be true **regardless** of the value of $X$.

But if we didn't correctly guess the functional form of linear component of the mean, $$\text{E}[Y|X=x] = \beta_0 + \beta_1 X_1 + ... + \beta_p X_p$$

Then the the residuals might have nonzero mean or nonconstant variance for some values of $x$.

#### Residuals versus fitted values

To look for these issues, we can plot the residuals $e_i$ against the fitted values $\hat y_i$:

```{r}
autoplot(bw_lm2, which = 1, ncol = 1) |> print()
```

If the model is correct, the blue line should stay flat and close to 0, and the cloud of dots should have the same vertical spread regardless of the fitted value.

If not, we probably need to change the functional form of linear component of the mean, $$\text{E}[Y|X=x] = \beta_0 + \beta_1 X_1 + ... + \beta_p X_p$$

#### Scale-location plot

We can also plot the square roots of the absolute values of the standardized residuals against the fitted values:

::: {.content-visible when-format="html"}
```{r}
autoplot(bw_lm2, which = 3, ncol = 1) |> print()
```
:::

::: {.content-visible when-format="pdf"}
```{r}
autoplot(bw_lm2, which = 3, ncol = 1) |> print()
```
:::

Here, the blue line doesn't need to be near 0, but it should be flat. If not, the residual variance $\sigma^2$ might not be constant, and we might need to transform our outcome $Y$ (or use a model that allows non-constant variance).

#### Residuals versus leverage

We can also plot our standardized residuals against "leverage", which roughly speaking is a measure of how unusual each $x_i$ value is. Very unusual $x_i$ values can have extreme effects on the model fit, so we might want ot remove those observations as outliers, particularly if they have large residuals.

::: {.content-visible when-format="html"}
```{r}
autoplot(bw_lm2, which = 5, ncol = 1) |> print()
```
:::

::: {.content-visible when-format="pdf"}
```{r}
autoplot(bw_lm2, which = 5, ncol = 1) |> print()
```
:::

The blue line should be relatively flat and close to 0 here.

### Diagnostics constructed by hand

```{r}

bw = 
  bw |> 
  mutate(
    predlm2 = predict(bw_lm2),
    residlm2 = weight - predlm2,
    std_resid = residlm2 / sigma(bw_lm2),
    # std_resid_builtin = rstandard(bw_lm2), # uses leverage
    sqrt_abs_std_resid = std_resid |> abs() |> sqrt()
    
  )

```

##### Residuals vs fitted

```{r}

resid_vs_fit = bw |> 
  ggplot(
    aes(x = predlm2, y = residlm2, col = sex, shape = sex)
  ) + 
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0)

```

::: {.content-visible when-format="html"}
```{r}
print(resid_vs_fit)
```
:::

::: {.content-visible when-format="pdf"}
```{r}
print(resid_vs_fit)
```
:::

##### Standardized residuals vs fitted

```{r}
bw |> 
  ggplot(
    aes(x = predlm2, y = std_resid, col = sex, shape = sex)
  ) + 
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0)

```

##### Standardized residuals vs gestational age

```{r}

bw |> 
  ggplot(
    aes(x = age, y = std_resid, col = sex, shape = sex)
  ) + 
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0)

```

##### `sqrt(abs(rstandard()))` vs fitted

Compare with `autoplot(bw_lm2, 3)`

```{r}


bw |> 
  ggplot(
    aes(x = predlm2, y = sqrt_abs_std_resid, col = sex, shape = sex)
  ) + 
  geom_point() +
  theme_classic() +
  geom_hline(yintercept = 0)


```

## Model selection

If we have a lot of covariates in our dataset, we might want to choose a small subset to use in our model.

There are a few possible metrics to consider for choosing a "best" model.

### Mean squared error

We might want to minimize the **mean squared error**, $\text E[(y-\hat y)^2]$, for new observations that weren't in our data set when we fit the model.

Unfortunately, $$\frac{1}{n}\sum_{i=1}^n (y_i-\hat y_i)^2$$ gives a biased estimate of $\text E[(y-\hat y)^2]$ for new data. If we want an unbiased estimate, we will have to be clever.

### Cross-validation

```{r}
data("carbohydrate", package = "dobson")
library(cvTools)
full.model <- lm(carbohydrate ~ ., data = carbohydrate)
temp = 
  cvFit(full.model, data = carbohydrate, K = 5, R = 10,
y = carbohydrate$carbohydrate)
```

## Categorical covariates with more than two levels

### Example: `birthweight`

In the birthweight example, the variable `sex` had only two observed values:

```{r}
unique(bw$sex)
```

If there are more than two observed values, we can't just use a single variable with 0s and 1s.

### 

For example, here's the (in)famous `iris` data:

```{r}

iris |> tibble()

summary(iris)

```

### 

There are three species:

```{r}

iris$Species |> unique()

```

### 

If we want to model `Sepal.Length` by species, we could create a variable $X$ that represents "setosa" as $X=1$, "virginica" as $X=2$, and "versicolor" as $X=3$.

```{r}
data(iris) # this step is not always necessary, but ensures you're starting  
# from the original version of a dataset stored in a loaded package

iris = 
  iris |> 
  tibble() |>
  mutate(
    X = case_when(
      Species == "setosa" ~ 1,
      Species == "virginica" ~ 2,
      Species == "versicolor" ~ 3
    )
  )

iris |> 
  distinct(Species, X) |> 
  print()

```

Then we could fit a model like:

```{r}

iris_lm1 = lm(Sepal.Length ~ X, data = iris)
iris_lm1 |> parameters() |> print_md()

```

### Let's see how that model looks:

```{r}

iris_plot1 = iris |> 
  ggplot(
    aes(
      x = X, 
      y = Sepal.Length)
  ) +
  geom_point(alpha = .1) +
  geom_abline(
    intercept = coef(iris_lm1)[1], 
    slope = coef(iris_lm1)[2]) +
  theme_bw(base_size = 18)
```

::: {.content-visible when-format="pdf"}
```{r}
print(iris_plot1)

```
:::

::: {.content-visible when-format="html"}
```{r}
ggplotly(iris_plot1)

```
:::

We have forced the model to use a straight line for the three estimated means. Maybe not a good idea?

### Let's see what R does with categorical variables by default:

```{r}
iris_lm2 = lm(Sepal.Length ~ Species, data = iris)
iris_lm2 |> parameters() |> print_md()

```

### Re-parametrize with no intercept

If you don't want the default and offset option, you can use "-1" like we've seen previously:

```{r}
iris.lm2b = lm(Sepal.Length ~ Species - 1, data = iris)
iris.lm2b |> parameters() |> print_md()

```

### Let's see what these new models look like:

```{r}
iris_plot2 = 
  iris |> 
  mutate(
    predlm2 = predict(iris_lm2)) |> 
  arrange(X) |> 
  ggplot(aes(x = X, y = Sepal.Length)) +
  geom_point(alpha = .1) +
  geom_line(aes(y = predlm2), col = "red") +
  geom_abline(
    intercept = coef(iris_lm1)[1], 
    slope = coef(iris_lm1)[2]) + 
  theme_bw(base_size = 18)
```

::: {.content-visible when-format="pdf"}
```{r}
print(iris_plot2)

```
:::

::: {.content-visible when-format="html"}
```{r}
ggplotly(iris_plot2)

```
:::

### Let's see how R did that:

```{r}

formula(iris_lm2)
model.matrix(iris_lm2) |> as_tibble() |> unique()

```

This is called a "corner point parametrization".

```{r}

formula(iris.lm2b)
model.matrix(iris.lm2b) |> as_tibble() |> unique()

```

This can be called a "group point parametrization".

There are more options; see Dobson & Barnett §6.4.1.
